MS draft

### 2020-05-04 doc/MS_v1.Rmd
commting during debugging
finializing edits

---
title: 'MS draft'
date: '`r Sys.Date()`'
output:
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    toc: yes
    toc_depth: '5'
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
always_allow_html: yes
bibliography: refs_Main_Report.bib
---

@@

library(raster)
library(lme4)
library(nlme)
library(RLRsim)
library(ggpubr)
library(reshape2)
library(patchwork)
library(png)
library(grid)
library(gridBase)
library(gridExtra)
library(dplyr)
library(cowplot)
library(kableExtra)

@@

<Mice were fed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison. Over the course of data collection some breeding colonies  mice were moved from facilities. Additionally the GI strain was kept at a separate facility.>

@@

The same dry-down spread technique was applied to both spermatocytes and oocytes based on [@peters_1997], with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used MLH1 quantification were from mice between 5 and 12 weeks, while juvillne mice ageed between 12 to 15 days were used for DMC1 quantification. Both pairs of ovaries were collected from embryo or neonate mice between 16 to 21 embyronic days and 0 to 48 hours old respctively.

Whole testes and decapsulated overaires were incubated in 3ml or 300ul of hypotonic solution for 45min respectively. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a photoflow wash (Kodak, cite). 
**slides were stored at -20*C if not stained imeediantly**

<Because some strains have breeding issues, all mice some strains were kept in breeding pairs (before collecting, and aged out of the desired age range. We performed tests for effects on MLH1 counts and found no effects in the older male mice ages. (Also we tested other effects, including maternal age...)>

The **immunohistochemistry**

@@

We tested for **effects** on mean MLH1 counts per cell; across mulitple mouse rooms, ages of mature males and differnces between embryos and neonates. No effects were found (Supplemental section).

### DMC1

Mean **DMC1** foci per cell for a single mouse from a subset of srains *musculus^PWD^*, *musculus^MSM^*, *domesticus^WSB^*, and *domesticus^G^* was quantified.

SC morphology and CREST/centromere number was used to stage spermatocytes **Certeria for early and late zygotene** - (bright staining - )
early zygotene defined as seperate centromere pairing -- mostly complete axis signal. late zygotene judged as -- low degree of centromere pairing -- and more homologous pairing (but not fully synapsis). Differences in DMC1 counts per cell were tested by ttests.
<numbers of cell?>

### MLH1 (Analytical Framework)

<Terminology Note, leptonema, zygonema, and pachynema diplonema diakinesis, are the stage names.   -ene endings refer to the cells/units within the cell stage? (I think)>

**certeria**  MLH1 foci were quantified from cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score, between 1 to 5, was assigned to each cell based on staining quality and general spread of bivalents. With scores of 1 and 2 having the highest quality. Cells and coded as a score of 5 were excluded from the final analysis. The distributions of MLH1 counts per cell was assessed for normality (supplemental figure). MLH1 foci located on the XY in spermatocytes were excluded from this study.

@@

Straight forward comparisons of groups (strain, sex) were done through t.tests().

We built the full mixed model (**M1**) to interpret the patterns of variation using the lmer() from the lmer4 package (cite) in R (v3.5.2)[@Rstudio]. Strain was coded as a random effect and tested using exactRLRT() to reflect a random sample of natural genetic variation from each subspecies. The main effects, subspecies, strain, and interaction were coded as fixed and tested by a chi-square test comparing the full and reduced models using drop1() and anova().

@@

### Single bivalent charaecterization

@@

The **total SC** per cell was quantified from the pachtyene cell images. To remove erroneous SC isolation, outliers were visually assessed at the mouse level and removed from the data set. Mouse means were calculated from cell-wide total SC lengths in `r length(  (MLH1.merge.Skel %>% filter(subsp != "Cast")  )$Random.Name)` out of `r length(  (MLH1_data %>% filter(subsp != "Cast")  )$Random.Name)` cells with MLH1 counts (Figure X).

The **DNA CrossOver** algorithm isolates single straightened bivalents/chromosomes shapes from an image, and returns SC length, and location of green and blue signal, reflecting MLH1 foci and cetromere signal respectively. This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this data set, isolation rates per cell (the number of all bivalents per cell) range from `r min(algorithm.stat.category$mean_pass.rate)` (molossinus^MSM^ male) to `r max(algorithm.stat.category$mean_pass.rate)` (musculus^KAZ^ female).

A curation step was applied to the total single bivalent data set to remove poor measures from the algorithm. From the total set of cell images, `r length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName )` bivalent objects were isolated by the image analysis software. After a human curation step , `r length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID )` single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step ([@peterson2019])).

From the curated single bivalent data set the **proportions** of bivalents by crossover number were quantified and tested by chi-square tests with the prop.test().

**REDUCED DATASET**

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. The 'short bivalent' data set included the four or five shortest bivalents, and excluded the X bivalent in oocytes. A total of 678 -- `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from 103 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and 37 -- `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes.  A 'long bivalent' data set was calculated as above the 4th quartille in SC lengths per cell.

#Results

```{r FOR.PLOT.organizing.data, echo=FALSE }
#set the order of another column, based on another variable. consistant order for ploting

@@

```{r Fig1.top, echo=FALSE, warning=FALSE}

@@

```{r female.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5, warning=FALSE}

@@

```{r male.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5, warning=FALSE}
#try customize the plot size

@@

```{r Figure1.Main.plot.three.plots, echo=FALSE, warning=TRUE, fig.width=7, fig.cap='*Figure 1 Mean MLH1 count distributions by strain.* A) Strain averages of MLH1 counts per cell. B) Female specific MLH1 count distributions. C)	Male specific MLH1 counts per cell.'}

@@

<notes about figure; make sure the order highlights the difference (mol, musc, dom). Clean up the axis and titles. Resizing might be an issue too.>

@@

To situate variation in recombination rate within an evolutionary framework, we fit a series of models including subspecies, strain, and sex, to mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice. We began with a full mixed model (M1, see Methods), which showed that strain 
(random effect p < <r MLH1.MMlmer_results\$rand.pval> `r 1/1000000`), sex (p = `r MLH1.MMlmer_results$sex.results`), subspecies (p=`r MLH1.MMlmer_results$sub.sp`), and subspecies sex (p = `r MLH1.MMlmer_results$interaction.pval`) each significantly affect recombination rate.

@@

We next fit general linear models separately for `r length(fomo.male_mouse.av$mouse)` males and `r length(female_mouse.avs_for.analysis$mouse)` females (M4; see methods). In the male dataset, three strains significantly affect recombination rate: _musculus^PWD^_ ((glm; p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]`; effect = 
`r sum_MLH1.male.glm.M4$coefficients[8]` foci), and 
_musculus^SKIVE^_ (glm; p = 0.01; `r sum_MLH1.male.glm.M4$coefficients[,4][10]`; effect = `r sum_MLH1.male.glm.M4$coefficients[10]`), and  _molossinus^MSM^_ (glm; p=`r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect  `r sum_MLH1.male.glm.M4$coefficients[9]`).

@@

<!-- transition gwRR to within mouse variance -->

@@

In general, females have almost twice as much variance in MLH1 foci per cell compared to males (Figure 1). Since estimates of within-mouse variance may be more susceptible to technical error from the staining protocol, we repeated the analyses using a subset of cells with higher quality scores (quality score 1 and 2, see Materials and Methods).  The results are similar: sex is the strongest effect (M1 p < `r 1/1000000` <round(LmerMLH1_M1_Q12_VAR_results\$sex.results, 5)>; M2 p = `r  round(sum.MLH1var_M2_Q12$coefficients[,4][4], 5)`;  also M3 p = `r MLH1var_M3_Q12_sum$coefficients[,4][2]`). When both quality-curated and full datasets are considered, strain does not significantly and consistently affect variance in MLH1 foci count per cell in either sex. These results suggest that within-mouse variance in recombination rate evolves independently of mean recombination rate.

**(sex specific results - no efficient strain effect for males. In females LEW - but this is due to cell quality difference, what about quality score and cell number within the models?**

<strain specific effect section was deleted for not highlighting main pattern. I agree> **re-read Lenzi et al -- heterogeneity in oogenesis**
<Bret's comment for below, are there differences between the curated and the full dataset? he propbably means, Q1 and full data set.>

## DSB. Evolution of genome-wide recombination rate is associated with evolution of double strand breaks

@@

#experimenting with ggboplot
everyday <- ggboxplot(DMC1.DF[DMC1.DF$stage != "P",], x = "strain", y = "total.count",
          color = "stage", palette = "jco",
          add = "jitter", short.panel.labs = FALSE)
  #library gpubr has functions to add in sig/test bars.

@@

  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.title.x = element_blank())

```{r Main.plot.DMC1, echo=FALSE, fig.width=6.5, fig.height=5, fig.cap="*Figure 2 male DSB estimates* A) Example early zygotene spermatocyte spread, B) Example late zygotene spermatocyte spread. Green foci show DMC1. B)	Distribution of DMC1 counts per cell  by strains. (1 mouse was used for each strain)"}

@@

three.DMC1.plot + plot_annotation(
  title = '',
  subtitle = '') + plot_annotation(tag_levels = 'A')

@@

anno_df_stage = compare_means(total.count ~ stage, group.by = "rec_group", data = DMC1.DF[DMC1.DF$stage != "P",],method = "t.test") %>% mutate(y_pos = 300)

@@

```{r DMC1_t.tests, echo=FALSE, warning=FALSE}
#2. t-tests for the DMC1 counts 
#need to account for sample sizes when comparing t.tests (KAZ and G don't have equal across )
#best case scnario -- more microscoping to get ~20 cells for each stage by strain combo

@@

In an attempt to localize the male-specific evolution of crossover number to steps of the meiotic pathway, we counted foci from a marker for double strand breaks (DSBs), DMC1, in prophase spermatocytes. DMC1 foci were scored from a total of `r sum( (DMC1table.main99$ncells) )` early zygotene-stage and `r sum( (DMC1table.main99$ncells) )` late zygotene-stage spermatocytes from three low-recombination strains ( _musculus^KAZ^_ , _domesticus^WSB^_ , and _domesticus^G^_ ) and two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_ ).

The high-recombination strains have significantly more DMC1 foci than the low-recombination strains in early zygotene cells (t-test, p < `r 1/1000000` <ttest.HighLow.L.pval>). In contrast, the two strain groups do not differ in DMC1 foci counted in late zygotene cells (t-test, p = `r ttest.HighLow.Z.pval`).

After DSB formation, DSBs are repaired as either non-crossovers (NCO) or as crossovers (COs), with the vast majority being repaired as NCOs. Thus the ratio of CO:DSB is a partial indicator of the proportion of DSBs which are designated as COs. The ratios, calculated for DMC1 means from both stages, are not significantly different between the high and low strain groups (t-test, p = `r DMC1.Lratio_t.test$p.value` and p = `r DMC1.Zratio_t.test$p.value` for early zygotene and late zygotene ratios, respectively). (connection to barier)

This comparison raises the possibility that the evolution of crossover number is primarily due to processes that precede the crossover/non-crossover decision. @cole2012 suggest that early zygotene counts for DMC1 are most relevant for predicting the final CO number due to the crossover homeostasis process.

This result, combined with the large difference in MLH1 count between high and low strains, might indicate that the CO/NCO decision contributes to strain differences in recombination rate.

<The number of DSBs increase in zygotene, then deminish as the effects of crossover homeostasis are implemented (Cole). (co homeostasis also indecates dsb ) Natural' variation in precursor number (the relationship of DSB and CO number) was analyzed in Barier et al  and links connects the CO number variation to 'pre' DSB formation fitting the model that DSB formation is dependant on the axis-loop structure of chromatin. (this study only examined one stage of early prophase for DMC1 counts)>

<the birer et al ratios for strain average DMC1: MLH1 were 5.6 cast. 6.6 C3H, 7.3 B6>

@@

To examine the connection between evolution of the genome-wide recombination rate and changes to the recombination landscape, we used an image analysis pipeline to measure properties of single bivalents ([@peterson2019]).

<bret suggests moving this section to Mat and Methods>
This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this dataset, isolation rates per cell range from `r min(algorithm.stat.category$mean_pass.rate)` ( _molossinus^MSM^_ male) to `r max(algorithm.stat.category$mean_pass.rate)` ( _musculus^KAZ^_ female).

From the total set of cell images, `r formatC(sum( length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName ) ), , format="d", big.mark="," )` bivalent objects were isolated by the image analysis software. After a human curation step ([@peterson2019]), `r formatC(sum(  length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID ) ), , format="d", big.mark=",")`  single-bivalent observations remained. We assume that the isolation of bivalents within cells is unbiased.

Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( total.biv.all, format="d", big.mark=",")`) have either one or two crossovers (Figure X). The proportion of one-crossover (1CO) to two-crossover (2CO) bivalents distinguishes the high vs. low recombining strains (Figure X). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` ( _musculus^SKIVE^_ ) in musculus^SKIVE^, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` in _musculus^PWD^_, and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` in _molossinus^MSM^_.

Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. _musculus^KAZ^_ p = `r PWD.KAZ.2CO.prop.test$p.value`; _molossinus^MSM^_ vs. _molossinus^MOLF^_ p = `r MSM.MOLF.2CO.prop.test$p.value`).

Next, we use this single-bivalent dataset to focus on aspects of the recombination landscape along chromosomes. We address two main questions. First, which traits are sexually dimorphic? Second, which traits differ between males from high-recombination vs. low-recombination strains?

## Q1. Sex Differences in the Length of the Synaptonemal Complex

@@

two.Q1.SC.plots + plot_annotation(
  title = '',
  subtitle = '') + plot_annotation(tag_levels = 'A')

@@

In many mammalian species, the synaptonemal complex (SC) is longer in females (@cahoonLibuda2019, @sardell_sex_2020). Yet, the majority of these observations come from species with female-biased heterochiasmy. Our wider survey of recombination in house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is male-biased. In addition, if SC length is a strong determinant of the genome-wide recombination rate,  male _musculus^PWD^_ and male _molossinus^MSM^_ should have longer SC lengths than the other strains. <moved section on reduced bivalent data set to methods>

<figure>

@@

## Q1. Sex Differences in the Positions of Single Crossovers

@@

plot1.WSB <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_WSB.png')
plot2.LEW <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_LEW.png')
plot3.G <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_G.png')

plot4.KAZ <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_KAZ.png')
plot5.SKIVE <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_SKIVE.png')
plot6.PWD <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_PWD.png')

plot7.MSM <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_MSM.png')

grid.arrange(rasterGrob(plot1.WSB), rasterGrob(plot2.LEW), rasterGrob(plot3.G),
             rasterGrob(plot4.KAZ), rasterGrob(plot5.SKIVE), rasterGrob(plot6.PWD),
             rasterGrob(plot7.MSM), 
             ncol=3)

#load all the files / arrange the plots from above 
#WSB.triangle.png <- readPNG("~./MLH1repo/doc/figureFiles/test.triangle_WSB.png")

@@

```{r Q1.single.biv.table, echo=FALSE}

#strain.BivData.Table

@@

<!-- plots -->

@@

```{r Q2.glm.SC.long, echo=FALSE}

#this dataset is all males
glmQ2_long.biv_M4 <- glm(mean.SC ~ subsp * strain, data= Q2mouse.av.long.biv.table)
sum.glmQ2_long.biv_M4 <- summary(glmQ2_long.biv_M4)

>>>>

---
title: 'MS draft'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  word_document:
  pdf_document:
    toc: yes
    toc_depth: '5'
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
always_allow_html: yes
bibliography: refs_Main_Report.bib
---

@@

library(raster)
library(lme4)
library(nlme)
library(RLRsim)
#library(ggpubr)
library(reshape2)
library(patchwork)
library(png)
library(grid)
library(gridBase)
library(gridExtra)
library(dplyr)
library(cowplot)
library(kableExtra)

@@

@@

The same dry-down spread technique was applied to both spermatocytes and oocytes based on [@peters_1997], with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used MLH1 quantification were from mice between 5 and 12 weeks, while juvillne mice ageed between 12 to 15 days were used for DMC1 quantification. Both pairs of ovaries were collected from embryo or neonate mice between 16 to 21 embyronic days and 0 to 48 hours old respctively. Whole testes and decapsulated overaires were incubated in 3ml or 300ul of hypotonic solution for 45min respectively. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, cite). Slides were stored at -20*C if not stained immediately.

@@

To estimate the numbers of percursor sites which may lead to crossovers, the mean **DMC1** foci per cell for a single mouse from a subset of srains *musculus^PWD^*, *musculus^MSM^*, *domesticus^WSB^*, and *domesticus^G^* was quantified. SC morphology and CREST/centromere foci number was used to stage spermatocytes, in early and late zygotema stages.

**MLH1** foci were quantified from cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score, between 1 to 5, was assigned to each cell based on staining quality and general spread of bivalents. With scores of 1 and 2 having the highest quality. Cells and coded as a score of 5 were excluded from the final analysis. The distributions of MLH1 counts per cell was assessed for normality (supplemental figure). MLH1 foci located on the XY in spermatocytes were excluded from this study.

@@

A series of mixed models and general linear models were used as an analetical framework. We built the full mixed model (**M1**) to interpret the patterns of variation using the lmer() from the lmer4 package (cite) in R (v3.5.2)[@Rstudio]. Strain was coded as a random effect and tested using exactRLRT() to reflect a random sample of natural genetic variation from each subspecies. The main effects, subspecies, strain, and interaction were coded as fixed and tested by a chi-square test comparing the full and reduced models using drop1() and anova().

@@

### Single bivalent characterization

@@

The **total SC** area per cell was quantified from the pachtyene cell images. To remove erroneous SC isolation, outliers were visually assessed at the mouse level and removed from the data set. Mouse means were calculated from cell-wide total SC lengths in `r formatC( length(  (MLH1.merge.Skel %>% filter(subsp != "Cast")  )$Random.Name), format="d", big.mark=",")` out of `r formatC( length(  (MLH1_data %>% filter(subsp != "Cast")  )$Random.Name), format="d", big.mark=",")` cells with MLH1 counts (Figure X).

The **DNA CrossOver** algorithm isolates single straightened bivalents/chromosomes shapes from an image, and returns SC length, and location of green and blue signal, reflecting MLH1 foci and cetromere signal respectively. This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this data set, isolation rates per cell (the number of all bivalents per cell) range from `r min(algorithm.stat.category$mean_pass.rate)` ( _molossinus^MSM^_ male) to `r max(algorithm.stat.category$mean_pass.rate)` ( _musculus^KAZ^_ female).

A curation step was applied to the total single bivalent data set to remove poor measures from the algorithm. From the total set of cell images, `r formatC( length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName ), format="d", big.mark=",")` bivalent objects were isolated by the image analysis software. After a human curation step , `r formatC( length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID ), format="d", big.mark=",")` single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step ([@peterson2019])). From the curated single bivalent data set the **proportions** of bivalents by crossover number were quantified and tested by chi-square tests with the prop.test().

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we considered a **reduced data** set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. The 'short bivalent' data set included the four or five shortest bivalents, and excluded the X bivalent in oocytes. A total of `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and `r sum(  (Q1.real.short_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes.  A 'long bivalent' data set was calculated as above the 4th quartille in SC lengths per cell.

# Results

```{r FOR.PLOT.organizing.data, echo=FALSE, warning=FALSE}
#set the order of another column, based on another variable. consistant order for ploting

@@

```{r Fig1.top, echo=FALSE, warning=FALSE, message=FALSE}

@@

```{r female.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5, warning=FALSE, message=FALSE}

@@

```{r male.show.plots, echo=FALSE, fig.width = 4, fig.height=2.5, warning=FALSE, message=FALSE}
#try customize the plot size

@@

```{r Figure1.Main.plot.three.plots, echo=FALSE, warning=FALSE, fig.width=7, fig.cap='*Figure 1 Mean MLH1 count distributions by strain.* A) Strain averages of MLH1 counts per cell. B) Female specific MLH1 count distributions. C)	Male specific MLH1 counts per cell.', message=FALSE}

@@

@@

To situate variation in recombination rate within an evolutionary framework, we fit a series of models including subspecies, strain, and sex, to mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice. We began with a full mixed model (M1, see Methods), which showed that strain 
(random effect p < `r 1/1000000`), sex (p = `r MLH1.MMlmer_results$sex.results`), subspecies (p=`r MLH1.MMlmer_results$sub.sp`), and subspecies sex (p = `r MLH1.MMlmer_results$interaction.pval`) each significantly affect recombination rate.

@@

We next fit general linear models separately for `r length(fomo.male_mouse.av$mouse)` males and `r length(female_mouse.avs_for.analysis$mouse)` females (M4; see methods). In the male dataset, three strains significantly affect recombination rate: _musculus^PWD^_ ((glm; p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]`; effect = 
`r sum_MLH1.male.glm.M4$coefficients[8]` foci), and 
_musculus^SKIVE^_ (glm; p = `r sum_MLH1.male.glm.M4$coefficients[,4][10]`; effect = `r sum_MLH1.male.glm.M4$coefficients[10]`), and  _molossinus^MSM^_ (glm; p=`r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect  `r sum_MLH1.male.glm.M4$coefficients[9]`).

@@

@@

In general, females have almost twice as much variance in MLH1 foci per cell compared to males (Figure 1). Since estimates of within-mouse variance may be more susceptible to technical error from the staining protocol, we repeated the analyses using a subset of cells with higher quality scores (quality score 1 and 2, see Materials and Methods).  The results are similar: sex is the strongest effect (M1 p < `r 1/1000000`; M2 p = `r  round(sum.MLH1var_M2_Q12$coefficients[,4][4], 5)`;  also M3 p = `r MLH1var_M3_Q12_sum$coefficients[,4][2]`). When both quality-curated and full datasets are considered, strain does not significantly and consistently affect variance in MLH1 foci count per cell in either sex. These results suggest that within-mouse variance in recombination rate evolves independently of mean recombination rate.

<LmerMLH1_M1_Q12_VAR_results> <Bret's comment for below, are there differences between the curated and the full dataset? -- Q1 and full data set>

## Evolution of genome-wide recombination rate is associated with evolution of double strand breaks

@@

@@

  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.title.x = element_blank())
```

```{r Main.plot.DMC1, echo=FALSE, fig.width=6.5, fig.height=5, fig.cap="*Figure 2 male DSB estimates* A) Example early zygotene spermatocyte spread, B) Example late zygotene spermatocyte spread. Green foci show DMC1. B)	Distribution of DMC1 counts per cell  by strains. (1 mouse was used for each strain)", eval=FALSE}

@@

three.DMC1.plot + plot_annotation(
  title = '',
  subtitle = '') + plot_annotation(tag_levels = 'A')
```

@@

#anno_df_stage = compare_means(total.count ~ stage, group.by = "rec_group", data = DMC1.DF[DMC1.DF$stage #!= "P",],method = "t.test") %>% mutate(y_pos = 300)

@@

```{r DMC1_t.tests, echo=FALSE, warning=FALSE}
#2. t-tests for the DMC1 counts

@@

In an attempt to localize the male-specific evolution of crossover number to steps of the meiotic pathway, we counted foci from a marker for double strand breaks (DSBs), DMC1, in prophase spermatocytes. DMC1 foci were scored from a total of `r sum( (DMC1table.main99$ncells) )` early zygotene-stage and `r sum( (DMC1table.main99$ncells) )` late zygotene-stage spermatocytes from three low-recombination strains ( _musculus^KAZ^_ , _domesticus^WSB^_ , and _domesticus^G^_ ) and two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_ ). The high-recombination strains have significantly more DMC1 foci than the low-recombination strains in early zygotene cells (t-test, p < `r 1/1000000`). In contrast, the two strain groups do not differ in DMC1 foci counted in late zygotene cells (t-test, p = `r ttest.HighLow.Z.pval`).

<ttest.HighLow.L.pval>

After DSB formation, DSBs are repaired as either non-crossovers (NCO) or as crossovers (COs), with the vast majority being repaired as NCOs. Thus the ratio of CO:DSB is a partial indicator of the proportion of DSBs which are designated as COs. The ratios, calculated for DMC1 means from both stages, are not significantly different between the high and low strain groups (t-test, p = `r DMC1.Lratio_t.test$p.value` and p = `r DMC1.Zratio_t.test$p.value` for early zygotene and late zygotene ratios, respectively). This comparison raises the possibility that the evolution of crossover number is primarily due to processes that precede the crossover/non-crossover decision. @cole2012 suggest that early zygotene counts for DMC1 are most relevant for predicting the final CO number due to the crossover homeostasis process.  This result, combined with the large difference in MLH1 count between high and low strains, might indicate that the CO/NCO decision contributes to strain differences in recombination rate.

<!-- The number of DSBs increase in zygotema, then deminish as the effects of crossover homeostasis are implemented (Cole). (co homeostasis also indecates dsb ) Natural' variation in precursor number (the relationship of DSB and CO number) was analyzed in Barier et al  and links connects the CO number variation to 'pre' DSB formation fitting the model that DSB formation is dependant on the axis-loop structure of chromatin. (this study only examined one stage of early prophase for DMC1 counts -->

<!-- the birer et al ratios for strain average DMC1: MLH1 were 5.6 cast. 6.6 C3H, 7.3 B6 -->

@@

To examine the connection between evolution of the genome-wide recombination rate and changes to the recombination landscape, we used an image analysis pipeline to measure properties of single bivalents ([@peterson2019]). This algorithm substantially speeds the accurate measurement of bivalents, but has the limitation that not all bivalents per cell can be isolated due to overlapping bivalents. In this dataset, isolation rates per cell range from `r min(algorithm.stat.category$mean_pass.rate)` ( _molossinus^MSM^_ male) to `r max(algorithm.stat.category$mean_pass.rate)` ( _musculus^KAZ^_ female). From the total set of cell images, `r formatC(sum( length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName ) ), , format="d", big.mark="," )` bivalent objects were isolated by the image analysis software. After a human curation step ([@peterson2019]), `r formatC(sum(  length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID ) ), , format="d", big.mark=",")`  single-bivalent observations remained. We assume that the isolation of bivalents within cells is unbiased. Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( total.biv.all, format="d", big.mark=",")`) have either one or two crossovers (Figure X). The proportion of one-crossover (1CO) to two-crossover (2CO) bivalents distinguishes the high vs. low recombining strains (Figure X). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` ( _musculus^SKIVE^_ ) in musculus^SKIVE^, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` in _musculus^PWD^_, and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` in _molossinus^MSM^_. Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. _musculus^KAZ^_ p = `r PWD.KAZ.2CO.prop.test$p.value`; _molossinus^MSM^_ vs. _molossinus^MOLF^_ p = `r MSM.MOLF.2CO.prop.test$p.value`). Next, we use this single-bivalent dataset to focus on aspects of the recombination landscape along chromosomes. We address two main questions. First, which traits are sexually dimorphic? Second, which traits differ between males from high-recombination vs. low-recombination strains?

## Sex Differences in Meiotic Traits

@@

#two.Q1.SC.plots + plot_annotation(
#  title = '',
#  subtitle = '') + plot_annotation(tag_levels = 'A')

@@

In many mammalian species, the synaptonemal complex (SC) is longer in females (@cahoonLibuda2019, @sardell_sex_2020). Yet, the majority of these observations come from species with female-biased heterochiasmy. Our wider survey of recombination in house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is male-biased. In addition, if SC length is a strong determinant of the genome-wide recombination rate,  male _musculus^PWD^_ and male _molossinus^MSM^_ should have longer SC lengths than the other strains.

@@

### Positions of Single Crossovers

@@

plot1.WSB <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_WSB.png')
plot2.LEW <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_LEW.png')
plot3.G <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_G.png')
plot4.KAZ <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_KAZ.png')
plot5.SKIVE <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_SKIVE.png')
plot6.PWD <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_PWD.png')
plot7.MSM <- readPNG('~/MLH1repo/doc/figureFiles/IFD.triangle_MSM.png')

# grid.arrange(rasterGrob(plot1.WSB), rasterGrob(plot2.LEW), rasterGrob(plot3.G),
#             rasterGrob(plot4.KAZ), rasterGrob(plot5.SKIVE), rasterGrob(plot6.PWD),
#             rasterGrob(plot7.MSM), 
#             ncol=3)

#load all the files / arrange the plots from above 
#WSB.triangle.png <- readPNG("~./MLH1repo/doc/figureFiles/test.triangle_WSB.png")
```

@@

@@

@@

```{r Q2.glm.SC.long, echo=FALSE}
#this dataset is all males
glmQ2_long.biv_M4 <- glm(mean.SC ~ subsp * strain, data= Q2mouse.av.long.biv.table)
sum.glmQ2_long.biv_M4 <- summary(glmQ2_long.biv_M4)
